---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---


 **Generating matrix of the rsem estimated_counts**

```{r error = FALSE}
# Loading results

library(tidyverse)

setwd("C:\\Users\\mapanavici\\RStudio\\R")

Ov4Carbopool3 <- read.delim("WTCHG_626197_201106_.genes.results", header = TRUE)

Ovcar4Parental1 <- read.delim("WTCHG_626197_289105_.genes.results", header = TRUE)

Ovcar4Parental2 <- read.delim("WTCHG_626197_290117_.genes.results", header = TRUE)

Ovcar4Parental3 <- read.delim("WTCHG_626197_291129_.genes.results", header = TRUE)

Ov4Carbopool1 <- read.delim("WTCHG_626197_295177_.genes.results", header = TRUE)
 
Ov4Carbopool2 <- read.delim("WTCHG_626197_296189_.genes.results", header = TRUE)

```



```{r}
# Obtaining "gene_id" and "expected_count" columns, renaming "expected_count" column to individual sample.names

get_counts <- function(sample_df) {
  
  id_and_counts <- select(sample_df, gene_id, expected_count)
  names(id_and_counts)[2] <- deparse(substitute(sample_df))
  
  return(id_and_counts)
}

df1 <- get_counts(Ovcar4Parental1)
df2 <- get_counts(Ovcar4Parental2)
df3 <- get_counts(Ovcar4Parental3)
df4 <- get_counts(Ov4Carbopool1)
df5 <- get_counts(Ov4Carbopool2)
df6 <- get_counts(Ov4Carbopool3)

```



```{r}
# Merging all data sets by "gene_id" column 

rawCounts <- list(df1, df2, df3, df4, df5 ,df6) %>% 
                reduce(inner_join, by='gene_id')

```


**Differential Expression Analysis**

```{r error = FALSE}
if (!require("BiocManager", quietly = TRUE)) 
  install.packages("BiocManager") 
  BiocManager::install("DESeq2")

library("DESeq2")

```



```{r}
# Tidying up the data for DESeq2

rownames(rawCounts) <- rawCounts$gene_id

rawCounts <-  select(rawCounts, -1)

rawCounts <- ceiling(rawCounts) # converting floats to int for DESeq2

head(rawCounts)

```



```{r}
# Inputting - Sample Groups

sampleData <- read.delim("sample_IDs.csv", header = T)

sampleData <- sampleData %>% 
      separate(sample.name.group, c("sample.name", "group"))

head(sampleData)
```



```{r}
# Creating DESeq2 object

dds <- DESeqDataSetFromMatrix(countData = rawCounts,
                              colData = sampleData,
                              design = ~ group)

dds
```


```{r}
# Filtering out low count genes - allowing 10 counts in at least 3 samples

keep <- rowSums(counts(dds) >= 10) >= 3

dds <- dds[keep,]

```



```{r}
# Running DESeq2 algo

dds <- DESeq(dds)

```


```{r}
# Using unsupervised clustering to see the relationship among two groups

# Data normalization

norm.counts <- counts(dds, normalized=TRUE)

write.csv(norm.counts, file="normalized_counts.txt", row.names=TRUE)

head(norm.counts)

# Data transformation (using variance stabilizing transformations (VST) which produces transformed data on the log2 scale, normalized with respect to library size)

vsd <- vst(dds, blind=FALSE)

head(assay(vsd))
```



```{r}
# Hierarchical clustering (dendrogram)

sampleDists <- dist(t(assay(vsd)))
plot(hclust(sampleDists, method = "complete"))

```



```{r}
# PCA

plotPCA(vsd, intgroup=c("group"))
```



```{r}
# Getting the results of differential expression analysis

res <- results(dds, contrast=c("group", "sensitive", "resistant"))

# Viewing summary of results

summary(res)
#head(res)
#mcols(res)$description
```


```{r}
# Diagnostics - observing how well the data has been captured by the model (plotting Dispersion Estimates)

plotDispEsts(dds)

```


```{r}
# Graphical Visualizations of the DE genes – MA plot (shows the log2 fold changes of the genes (each dot is a gene) over the mean of normalized counts for all samples in our my data)

plotMA(res)

```
Positive log2 fold changes values represent upregulated genes in sensitive compared to resistant group 
Negative log2 fold changes values represent dowregulated genes in sensitive compared to resistant group 


```{r}

# Graphical Visualizations of the differentially expressed genes, subseting the DE genes with padj < 0.05 and log2FC > |1| i.e. at least two-fold change in expression

library(pheatmap)

de.sign <- subset(res, padj < 0.05 & abs(log2FoldChange) >1 )

de.sign.genes <- rownames(de.sign)

scale.dat <- t(scale(t(assay(vsd)[de.sign.genes,])))

pheatmap(scale.dat[de.sign.genes,], cluster_rows=TRUE, show_rownames=FALSE, cluster_cols=TRUE)

```


```{r}
# Annotation of output tables 

BiocManager::install('AnnotationDbi')
library(AnnotationDbi)
library(org.Hs.eg.db)


res.df <- as.data.frame(res)

res.annot <- mapIds(org.Hs.eg.db, keys = rownames(res.df), keytype = "ENSEMBL", column = "SYMBOL")

res.df.annot <- cbind(res.df,gene.symbol = res.annot) #binding the annotations to the sign.df

head(res.df.annot)

write.csv(res.df.annot, file = "DE_sensitiveVSresistant_results_wGeneSymbol.csv")

```


```{r}

# Using volcano plot to visualize the Log2Fold changes of all genes (again each dot represents a gene) plotted against the -Log10 Adjusted pvalue.
# In red are genes with padj < 0.05 and log2fold change more or less than 1. They are either differently upregulated (right) or downregulated (left) in sensitive samples compare to resistant.

BiocManager::install('EnhancedVolcano')
library(EnhancedVolcano)


EnhancedVolcano(res.df.annot,
    lab = res.df.annot$gene.symbol,
    x = 'log2FoldChange',
    y = 'padj')

```


**Gene set enrichment analysis**

```{r}
# Performing Pre-Ranked Gene Set Enrichment Analysis (GSEA Pre-Ranked) to understand better what biological processes and pathways are affected 

library(fgsea)

library(data.table)

library(ggplot2)

library(qusage)
```



```{r}
# Loading DE results

DE.res <- read.csv("DE_sensitiveVSresistant_results_wGeneSymbol.csv", header = TRUE, row.names = 1)
DE.res
```


```{r}
# Loading the Gene Ontology Collection

gmt.file <- read.gmt("c5.go.bp.v7.4.symbols.gmt")
```



```{r}
# Ranking DE results by decreasing Log2 Fold-change and then creating a named vector of the log2 fold changes with the names being the gene symbols.

DE.res.ranked <- DE.res[order(DE.res$log2FoldChange, decreasing = T), ]

DE.ranks <- setNames(DE.res.ranked$log2FoldChange, DE.res.ranked$gene.symbol)
```



```{r}
# Running the fgsea function

fgseaRes <- fgsea(gmt.file, DE.ranks, minSize=15, maxSize=500)
head(fgseaRes)

```


```{r}
# ordered from most sign. to least 

topPathways <- fgseaRes[order(padj)]

topPathways
```


**Pathway Enrichment Analysis**

```{r}
# selecting sign. DE genes

library("AnnotationDbi")
library("org.Hs.eg.db")


# adding symbol column for further pathway analysis

res$symbol = mapIds(org.Hs.eg.db,
                    keys=row.names(res), 
                    column="SYMBOL",
                    keytype="ENSEMBL",
                    multiVals="first")

sig_genes <- res[res$padj <= 0.05 & !is.na(res$padj), "symbol"]

print(paste("Total number of significant genes:", length(sig_genes)))

```


```{r}
write.table(sig_genes, file="significant_genes.txt", row.names=FALSE, col.names=FALSE, quote=FALSE)
```


Performing pathway analysis online where I uploaded "significant_genes.txt" to Reactome website (https://reactome.org/PathwayBrowser/#TOOL=AT). Selectied parameters:“Project to Humans”
```{r}
reactome_res <- read.csv("reactome_result.csv")
reactome_res
```


```{r}
# top dysregulated pathways

reactome_res %>% dplyr::select(2, 6)
```
**Gene Ontology Biological Pathways (GOBP) related to the tumor micro-environment, particularly the extracellular matrix, were enriched in carboplatin-resistant cells**


